Formalization of Typed Lua

### 2015-06-12 doc/papers/paper/paper.tex
DLS paper: simplify rule T-CONSTRUCTOR-FIELDS
DLS: put examples in introduction to make motivation clearer

In this paper, we give a formal basis to an updated version of that design, with a focus on novel type system features
that Typed Lua uses to type some tricky Lua idioms.
One of these idioms is the use of imperative updates to
build records and objects one field at a time; Typed Lua
lets record and object types evolve incrementally along
with these updates while not sacrificing width subtyping.
Another idiom combines overloading on the return values
of a function, multiple return values, multiple assignment,
and tag checks to discriminate a several returned values
by checking just one of them; Typed Lua deconstructs unions
of tuples in a way that the dependencies between the types
of each value in the tuple can be tracked.

@@

The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-FIELDS}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

@@

As an example, consider the integer division function
below, which returns the quotient and rest of integer
division, or a division by zero error:
\begin{verbatim}
  local function idiv(x : integer, y : integer)
    if divisor == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end
\end{verbatim}

Typed Lua infers the return type of this function as
$(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such
a function is to test the first returned value:
\begin{verbatim}
  local q, r = idiv(a, b)
  if q then
    -- assume q and r are integers
  else 
    -- assume q is nil r is a string
  end
\end{verbatim}

>>>>

In this paper, we give a formal basis to an updated version of that design, with a focus on novel type system features
that Typed Lua uses to type some tricky Lua idioms.
One of these idioms is the use of assignment to
build records and objects one field at a time, as in
the following example, where the {\tt newshape} function
is a constructor for objects with {\tt x} and {\tt y}
fields and a {\tt move} method:
\begin{verbatim}
  function new_shape(x, y)
    local shape = {}
    shape.x = x
    shape.y = y
    function shape:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return shape
  end
\end{verbatim}

The constructor function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

Another idiom combines overloading on the return values
of a function, multiple return values, multiple assignment
to decompose multiple return values,
and tag checks to discriminate a several returned values
by checking just one of them, as in the following example:
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}

In the {\tt then} branch it is guaranteed that both {\tt q}
and {\tt r} are integers, and in the {\bf else} branch it is
guaranteed that {\tt q} is {\tt nil} and {\tt r} is a string.
In Typed Lua, the return type of {\tt idiv} can be a union of
tuples describing the possible return values. Typed Lua deconstructs unions of tuples in a way that the dependencies between the types of each value in the tuple can be tracked.

@@

The rule \textsc{T-CONSTRUCTOR-FIELDS} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR-FIELDS}\\
\dfrac{\begin{array}{c}
       \overline{\env_1, \penv \vdash e_2: V, \env_v}
        \;\;\;
       \overline{\env_v, \penv \vdash e_1: K, \env_f}
        \\
       T = \{\overline{K{:}V}\}_{unique} \;\;\;
       \env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
%\dfrac{\begin{array}{c}
%	\env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
%	T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
%	wf(T) \;\;\;
%	n = |\;\overline{[e_{1}] = e_{2}}\;| \\
%	\env_{f} = merge(\env_{1}, ..., \env_{n+1})
%	\end{array}}
%{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
%\end{array}
\]

@@

As an example, consider again the integer division function
from Section~1, which returns the quotient and rest of integer
division, or a division by zero error, now statically typed:
\begin{verbatim}
  local function idiv(x : integer, y : integer):
        (integer, integer) | (nil, string)
    if divisor == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end
\end{verbatim}

The return type of this function is
$(\Integer \times \Integer \times \Nil*) \sqcup (\Nil \times \String \times \Nil{*})$. The idiom for using this such
a function is to test the first returned value:
\begin{verbatim}
  local q, r = idiv(a, b)
  if q then
    -- assume q and r are integers
  else 
    -- assume q is nil r is a string
  end
\end{verbatim}
